---
title: "GWAS Manhattan plot example"
author: "Mikołaj Buchwald"
date: "Created on 15 Apr, 2021. Last compiled on `r format(Sys.time(), '%d %b, %Y')`."
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Manhattan plot of human genome

## Introduction

Manhattan plot is used to visualize large datasets, with large number of data points, and a significant distribution (variance) for some of these points (features). This is a primary visualization tool in genom-wide association studies (GWAS).

For a basic example of a Manhattan plot used in genomics reasearch see: https://en.wikipedia.org/wiki/Manhattan_plot

The below example is heavely based on the Vignette (documentation, example) from the R Bioconducotr `Sushi` package by Douglas H. Phanstiel, Alan P. Boyle, Carlos L. Araya, and Mike Snyder (https://www.bioconductor.org/packages/release/bioc/vignettes/Sushi/inst/doc/Sushi.pdf)

***

## Install `Sushi` package

In oder to produce a Manhattan plot with R Bioconductor, one can use the `Sushi` package (https://www.bioconductor.org/packages/release/bioc/vignettes/Sushi/inst/doc/Sushi.pdf):

```{r sushi}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Sushi")

library('Sushi')
```

## Data

### Chromosomes

`Sushu_hg18_genome` stores the information on the length of human chromosomes as described by NCBI36/hg18 genome build (see: https://www.bioconductor.org/packages/release/bioc/manuals/Sushi/man/Sushi.pdf).

We can display the number of base pairs in each chromosome (in millions, i.e., 1*10^6^):

```{r base_pairs}
data(Sushi_hg18_genome)
data.frame(Sushi_hg18_genome$V1, Sushi_hg18_genome$V2/1000000)
```

For the reference number of base pairs per chromosome see: https://www.ncbi.nlm.nih.gov/books/NBK22266/.

### GWAS `bed` data

Bed data representing human genes with coordinates from NCBI36/hg18 genome build (see: https://www.bioconductor.org/packages/release/bioc/manuals/Sushi/man/Sushi.pdf).

The beginning of the GWAS bed data frame:

```{r gwas_bed}
data(Sushi_GWAS.bed)
head(Sushi_GWAS.bed)
```

Number of coordinates (number of rows):

```{r nrows}
nrow(Sushi_GWAS.bed)
```

Note the names of the columns:
```{r colnames}
names(Sushi_GWAS.bed)
```

## Manhattan Plot

A "genome" parameter is required to set how many base pairs there are in each chromosome, which also is represented in the graph in how many space particular chromosome takes.

Comment: `Sushi_GWAS.bed[,5]` is the `pval.GC.DBP` column

```{r pressure}
plotManhattan(bedfile=Sushi_GWAS.bed,
              pvalues=Sushi_GWAS.bed[,5],
              col=SushiColors(6),
              genome=Sushi_hg18_genome,
              cex=0.75)

labelgenome(genome=Sushi_hg18_genome,n=4,scale="Mb",
           edgeblankfraction=0.20,cex.axis=.5)

axis(side=2,las=2,tcl=.2)
mtext("log10(P)",side=2,line=1.75,cex=1,font=2)
mtext("chromosome",side=1,line=1.75,cex=1,font=2)
```